Aligning sequence reads, clone sequences and assembly contigs with BWA-MEM

### 2013-03-14 bwamem.tex
improve wording, a little bit

\DeclareMathOperator*{\argmax}{argmax}

@@

\begin{methods}

@@

\subsection{Aligning single query sequence}

\subsubsection{Seeding and re-seeding} BWA-MEM follows the canonical
seed-and-extend paradigm. It initially seeds an alignment with supermaximal
exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm 5]{Li:2012fk}. This algorithm
essentially finds at each query position the longest exact match covering the
position. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
reference genome. If $l$ is too large (over 28bp by default), we then re-seed
with exact matches that cover the middle base of the SMEM and occur at least
$k+1$ times in the genome, which can be done by requiring a minimum occurrence
in the original SMEM algorithm. Re-seeding reduces the chance of missing a seed
from the true alignment.

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. While seeding, we chain
the seeds with a naive algorithm. We then filter out short chains that are
largely contained in a long chain and is much worse than the long chain (by
default, both 50\% and 38bp shorter than the long chain). Note that chain
filtering aims to reduce unsuccessful seed extension at a later step. The
chains detected here do not need to be accurate. We may implicitly split or
merge chains during seed extension.

\subsubsection{Seed extension} We rank the seeds by the length of the chains
the seeds belong to and then by the length of the seeds in each chain. We start
from the highest ranked seed and for each seed, we drop the seed if it is
already contained in an alignment found before, or extend with a
affine-gap-penalty dynamic programming (DP) otherwise. We will find all the
potential alignments of a query sequence after this step.

BWA-MEM's seed extension differs from the standard seed extension in two
aspects. Firstly, suppose at a certain extension step we come to reference
position $x$ with the best extension score achieved at query position $y$.
BWA-MEM will stop extension if the score at $(x,y)$ is smaller than the best
extension score observed so far by $Z+|x-y|\times p_{\rm gapExt}$, where
$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary value. This heuristic avoids extending
through a poorly aligned region with good flanking alignment. It is similar
to the X-dropoff heuristic in BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in
one of the reference or the query sequences.

Secondly, while extending a seed, BWA-MEM tries to keep track of the best
extension score reaching the end of the query sequence. If the difference
between the best score reaching the end and the best local alignment score is
below a threshold, which is called clipping penalty, the local alignment will
be rejected even if it has a higher score. BWA-MEM uses this strategy to
automatically choose between local and end-to-end alignments. It may allow true
substitutions towards the end of a read and thus reduces reference bias, while
avoids introducing excessive mismatches and gaps which may happen when we force
an end-to-end alignment for a read bridging a break point associated with a
long insertion or deletion (indel) or a structural variation or a misassembly.

@@

\subsubsection{Estimating insert size distribution and rescuing missing hits}
Like BWA, BWA-MEM processes a batch of reads at a time.  Given paired-end data,
it estimates the mean $\mu$ and the variance $\sigma^2$ of the insert size
distribution from reliable single-end alignments. For the top 100 hits (by
default) of each end, BWA-MEM performs the Smith-Waterman alignment (SW) for
its mate within a window $[\mu-4\sigma,\mu+4\sigma]$ from the hit. The second
best alignment score is recorded to detect potential mismapping in a long
tandem repeat.  Hits found both from the single-sequence alignment and SW
rescuing will be fed to the next step for pairing.

\subsubsection{Pairing} Given the $i$-th hit for the first read and $j$-th hit
for the second, BWA-MEM computes their distance $d_{ij}$ if the two hits are in
the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
pair $(i,j)$ as $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$ and
$S_j$ are the SW score of the two reads in the pair, respectively, $a$ is the
matching score and $P(d)$ gives the probability of observing an insert size
larger than $d$ assuming a normal distribution. `$\log_4$' arises when we
relate SW score to a probabilistic model. $U$ is the chimeric penalty,
which controls preference of pairing: if $d_{ij}$ is small enough such
that $-a\log_4 P(d_{ij})<U$, BWA-MEM prefers to pair the two ends; otherwise
it prefers the unpaired alignments. BWA-MEM chooses the pair $(i,j)$ that
maximizes $S_{ij}$ as the final alignment for both ends.
This pairing strategy jointly considers the alignment scores of both ends,
the insert size and the possibility of chimeric read pairs.

@@

We evaluated the performance of BWA-MEM on simulated data together with
GEM, Bowtie2, Cushaw2, BWA-SW, BWA~\citep{Li:2009uq} and NovoAlign
(http://novocraft.com) (Figure~\ref{fig:eval}). On accuracy, NovoAlign
is the most accurate. BWA-MEM is close to NovoAlign for PE reads
and is comparable to GEM and Cushaw2 for SE. On speed, BWA-MEM is similar to
GEM and Bowtie2 for this data set, but is about 6 times as fast as Bowtie2
for a 650bp long-read data set.

To evaluate the performance for even long query sequences, we aligned the {\it E. coli}
K-12 MG1655 substrain (AC:NC\_000913) against the 536 strain (AC:NC\_008253) with 
both BWA-MEM and nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds, giving
105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of which 102,241
overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short regions
with very high divergence. It is unclear which aligner is better. Although
Nucmer is faster in the evaluation, only BWA-MEM scales to mammalian genomes.

BWA-MEM is a fast and accurate alignment algorithm for sequence reads and is
one of the few aligners that work efficiently with both short reads and long sequences up
to a few megabases. It is still possible to greatly accelerate BWA-MEM for long
sequences by using SSE2-powered banded DP and by restricting DP to regions not
covered by a long seed.

>>>>

@@

\vspace*{-1em}
\begin{methods}

@@

\subsection{Aligning a single query sequence}

\subsubsection{Seeding and re-seeding} BWA-MEM follows the canonical
seed-and-extend paradigm. It initially seeds an alignment with supermaximal
exact matches (SMEMs) using an algorithm we found previously~\citep[Algorithm
5]{Li:2012fk}, which essentially finds at each query position the longest exact
match covering the position. However, occasionally the true alignment may not
contain any SMEMs.  To reduce mismappings caused by missing seeds, we introduce
re-seeding. Suppose we have a SMEM of length $l$ with $k$ occurrences in the
reference genome. If $l$ is too large (over 28bp by default), we then re-seed
with exact matches that cover the middle base of the SMEM and occur at least
$k+1$ times in the genome. Such seeds can be found by requiring a minimum
occurrence in the original SMEM algorithm.

\subsubsection{Chaining and chain filtering} We call a group of seeds that are
colinear and close to each other as a \emph{chain}. We chain the seeds while
seeding. After that, we filter out short chains that are largely contained in a
long chain and are much worse than the long chain (by default, both 50\% and
38bp shorter than the long chain). Note that chain filtering aims to reduce
unsuccessful seed extension at a later step. Each chain may not always
correspond to a final hit. The chains detected here do not need to be accurate.

\subsubsection{Seed extension} We rank a seed by the length of the chain it
belongs to and then by the seed length in the chain. For each seed in the
ranked list, from the best to the worst, we drop the seed if it is already
contained in an alignment found before, or extend the seed with a banded
affine-gap-penalty dynamic programming (DP) if it potentially leads to a new
alignment.

BWA-MEM's seed extension differs from the standard seed extension in two
aspects. Firstly, suppose at a certain extension step we come to reference
position $x$ with the best extension score achieved at query position $y$.
BWA-MEM will stop extension if the difference between the best extension score
and the score at $(x,y)$ is larger than $Z+|x-y|\times p_{\rm gapExt}$, where
$p_{\rm gapExt}$ is the gap extension penalty and $Z$ is an arbitrary cutoff.
This heuristic avoids extension through a poorly aligned region with good
flanking alignment. It is similar to the X-dropoff heuristic in
BLAST~\citep{Altschul:1997vn}, but does not penalize long gaps in one of the
reference or the query sequences.

Secondly, while extending a seed, BWA-MEM tries to keep track of the best
extension score reaching the end of the query sequence. If the difference
between the best score reaching the end and the best local alignment score is
below a threshold, the local alignment will be rejected even if it has a higher
score. BWA-MEM uses this strategy to automatically choose between local and
end-to-end alignments. It may allow true substitutions towards the end of a
read and thus reduces reference bias, while avoids introducing excessive
mismatches and gaps which may happen when we force an end-to-end alignment for
a read bridging a break point associated with a long insertion or deletion
(indel) or a structural variation or a misassembly.

@@

\subsubsection{Rescuing missing hits}
Like BWA~\citep{Li:2009uq}, BWA-MEM processes a batch of reads at a time.
For each batch, it estimates the mean $\mu$ and the variance $\sigma^2$
of the insert size distribution from reliable single-end hits. For the
top 100 hits (by default) of each end, BWA-MEM performs SSE2-based
Smith-Waterman alignment (SW; \citealt{Farrar:2007hs}) for the mate within a
window $[\mu-4\sigma,\mu+4\sigma]$ from each hit. The second best SW score is
recorded to detect potential mismapping in a long tandem repeat. Hits found
from both the single-sequence alignment and SW rescuing will be used for
pairing. It is worth noting that BWA and BWA-SW only performs SW for the
best unique hit, but BWA-MEM also does SW for repetitive hits.

\subsubsection{Pairing} Given the $i$-th hit for the first read and $j$-th hit
for the second, BWA-MEM computes their distance $d_{ij}$ if the two hits are in
the right orientation, or sets $d_{ij}$ to infinity otherwise. It scores the
pair $(i,j)$ with $S_{ij}=S_i+S_j-\min\{-a\log_4 P(d_{ij}),U\}$, where $S_i$
and $S_j$ are the SW score of the two hits, respectively, $a$ is the matching
score and $P(d)$ gives the probability of observing an insert size larger than
$d$ assuming a normal distribution. `$\log_4$' arises when we interpret SW
score as odds ratio~\citep{Durbin:1998uq}. $U$ is a threshold that controls
pairing: if $d_{ij}$ is small enough such that $-a\log_4 P(d_{ij})<U$, BWA-MEM
prefers to pair the two ends; otherwise it prefers the unpaired alignments.
BWA-MEM chooses the pair $(i,j)$ that maximizes $S_{ij}$ as the final
alignments for both ends.  This pairing strategy jointly considers the
alignment scores, the insert size and the possibility of chimeric read pairs.

@@

We evaluated the performance of BWA-MEM on simulated data together with
NovoAlign (http://novocraft.com), GEM, Bowtie2, Cushaw2, BWA-SW and BWA
(Figure~\ref{fig:eval}). On accuracy, NovoAlign is the best. BWA-MEM is close
to NovoAlign for PE reads and is comparable to GEM and Cushaw2 for SE.  On
speed, BWA-MEM is similar to GEM and Bowtie2 for this data set, but is about
6 times as fast as Bowtie2 for a 650bp long-read data set.

To evaluate the performance for even longer query sequences, we aligned the
{\it E. coli} K-12 MG1655 substrain (AC:NC\_000913; 4.6Mb in length) against
the 536 strain (AC:NC\_008253) with both BWA-MEM and
nucmer~\citep{Kurtz:2004zr}. Nucmer finished the alignment in 25 seconds,
giving 105,505 substitution differences between the strains. BWA-MEM is slower.
It finished the alignment in 131 seconds, reporting 104,321 substitutions of
which 102,241 overlapping with the nucmer output. Manual examination of the
substitutions unique to each aligner reveals that most of them fall in short
regions with very high divergence. It is unclear which aligner is better. Note
that although Nucmer is faster in the evaluation, only BWA-MEM scales well to
large genomes.

BWA-MEM is a fast and accurate aligner for sequence reads and is one of the few
that work well for both short reads and long sequences up to a few megabases.
Technically, it is possible to make BWA-MEM even faster for long sequences by
using SSE2-based banded DP and by restricting DP to regions not covered by a
long exact match.
